Characterising cheque dishonour cases in India: Causes for delays and policy implications

### 2022-04-14 DOC/introduction.tex
Fixing some errors

India has a slow judiciary --- courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. Slow judiciaries have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, minimising unnecessary judicial delays could help improve enforcement and enhance the overall rule of law. It is estimated that judicial delays cost India around 1.5\% of its GDP annually.\footcite{dey2016_cost} Against this backdrop, it is hardly surprising that tackling judicial delay has increasingly become a top priority for Indian judges and policymakers.

Link number of cases to State GDP. No of cases is directly linked to the economic activity. If delay hinders performance, then it will hurt the well performing states more. State GDP will suffer and it will have asymmetrical impact on Indian GDP. Insert the figure of SGDP and no of NI 138 cases.

One reason for the large burden on courts is believed to be the large share of \gls{ni} cases.\footnote{In particular, \S~138 of the Negotiable Instruments Act (Dishonour of cheque for insufficiency, etc., of funds in the account). Act 26 of 1881} As per the \gls{lci}, they represent 6.5\% and 7.8\% of all institutions and pendency in Indian courts, respectively \autocite{lci2014_arrears}. As per one order of the Supreme Court of India, they reflect more than 15\% of all criminal cases in the District Courts \autocite{sc2020_makwanavstate}. As per another order, they constitute 30\% of the total pendency in courts.\footcite[Similarly, a study published by the Department of Justice briefly touches on the burden of such cases on the judiciary and posits that they constitute 34\% of pending criminal cases in Maharashtra.][]{sc2020_138, mahadik2018_maharashtra}

Such knowledge is based on the assumption that cheque dishonour cases are a large burden on the judiciary and has thus focused on expediting disposal. For example, \gls{lci} (2008), by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts \autocite{lci2008_138, bhan2015_placing}. However, it did not define \textit{fast-track courts} or give guidance concerning how and where they would operate. Developing this work, DAKSH (2017) examined the behaviour of cheque dishonour cases in the Indian courts by looking at over 67000 cases. In most cases, it found that resolution is delayed well beyond statutorily prescribed timelines and that certain banks and financial institutions are frequent complainants \autocite{sridhar2017_cheque}. As per our own estimates, less than 20\% of the cases get disposed within the stimulated timeline (see Figure \ref{fig:stateSurvival} for comparison across states in India).

\begin{figure}[h]
 \centering
 \caption{Survival probability of cases across States}\label{fig:stateSurvival}
 \includegraphics[width = 0.9\textwidth]{surv_states-1.png}
\end{figure}

To this end, in 2020, the Supreme Court of India took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881” \autocite{sc2020_138}. Among other directions, the court (i) appointed a \gls{coe}\footnote{Headed by Hon’ble Mr Justice RC Chavan, former Judge of the Bombay High Court.} and (ii) appointed Amici Curiae to assist the court,\footnote{Mr Sidharth Luthra (Sr Advocate) and Mr K Parameshwar (Advocate).} to study processes to expedite disposal of complaints under \S~138 of the \gls{ni}. This is in a series of interventions targeted to understand cheque dishonour cases in India. As \cref{sec:history} shows, most knowledge and development concerning such cases in India has come from government institutions, i.e. the \gls{lci} and the judiciary. The Amici appointed by the Supreme Court also made recommendations to expedite the disposal of cases. This included (i) increasing the use of pre- and post-summons mediation, (ii) expediting the service of summons to reduce absconsion, (iii) addressing the multiplicity of proceedings, etc.\footcite[For details see ]{amicus2020_submission}

\textbf{What we do?}: This study, based on orders of the Supreme Court and the report of the Amici Curiae, attempts to assess the effect of suggested interventions on the expeditious trial of cases under \S~138 of the \gls{ni}. First, we empirically estimate the volume of NI 138 cases differing courts. Given the disagreement, any acceptance among policymakers that \gls{ni} cases (\textit{cheque dishonour cases} hereinafter) are the reason for India's slow judiciary is likely to lead to misguided solutions. The misrepresentation could lead to diverting resources from other routes of judicial reform. Thus, one requires a more detailed analysis of the proportion of such cases, their causes, timelines, etc. This information can help judges and policymakers better target interventions.\footcite[For the importance of accurate judicial data, see][]{damle2020_ecourtsData, daksh2020_data, damle2020_land}

Second, we examines the determinants of case duration as identified by the Amici Curiae and the Supreme Court.

At last, we evaluate the effect of 2015 Amendments on the duration of cases.

\textbf{Key findings/Summary of our results}

\textbf{How we contribute to the debate?}

The rest of this paper is organised as follows: after this introduction, \cref{sec:history} elaborates the history of cheque dishonour provisions in India and their relation with judicial delays. In \cref{sec:select-case-char}, based on the recommendations of the Amici Curiae and the Supreme Court, we figure out the key case characteristics that can be targeted by direct court interventions. \Cref{sec:methodology} describes the methodology, and \cref{sec:results} presents the results thereof. Finally, \cref{sec:conclusion} concludes the paper and presents the way forward.

>>>>

India has a slow judiciary - courts are clogged with large backlogs \autocite{moog1992delays, debroy2008justice, dutta2019modernise}. As of 8th December 2021, over forty-one million cases were pending across district courts \autocite{njdg2021}. It is estimated that judicial delays cost India around 1.5\% of its GDP annually \autocite{dey2016_cost}. Slow judiciaries also have adverse consequences on the structure and efficiency of markets and the quality of life of citizens \autocite{world2004world, chemin2007impact, rao2020institutional}. Therefore, it is hardly surprising that tackling judicial delay has become a top priority for Indian judges and policymakers.

One reason for the large burden on courts is believed to be the large share of \gls{ni} cases.\footnote{In particular, \S~138 of the Negotiable Instruments Act (Dishonour of a cheque for insufficiency, etc., of funds in the account). Act 26 of 1881} While the NI Act contains several provisions that may lead to a dispute, \S~138, in particular, provides that if a cheque (drawn by a person for paying off a debt or a liability) is returned unpaid due to insufficiency of funds or credit, the payer may be imprisoned for up to two years or with a fine up to twice the amount of the cheque, or both. As per the \gls{lci}, they represent 6.5\% and 7.8\% of all institutions and pendency in Indian courts, respectively \autocite{lci2014_arrears}. As per one order of the Supreme Court of India, they reflect more than 15\% of all criminal cases in the District Courts \autocite{sc2020_makwanavstate}. As per another order, they constitute over 30\% of the total pendency in courts.\footcite[Similarly, a study published by the Department of Justice briefly touches on the burden of such cases on the judiciary and posits that they constitute 34\% of pending criminal cases in Maharashtra.][]{sc2020_138, mahadik2018_maharashtra}

Given this increasing volume and persistent mention of \gls{ni} cases as a major cause of judicial delay, \cite{sridhar2017_cheque} examined the behaviour of cheque dishonour cases in India by looking at the DAKSH database of 67,000 cases spread across Indian courts. In most cases, she found that resolution is delayed well beyond the statutorily prescribed timeline of 180 days. Using 50,000 case level data for eight States and two Union Territories, we also estimate the survival probability (using non-parametric Kaplan-Meier estimation) to understand the proportion of unresolved cases within the stipulated time.\footnote{The dataset used for this analysis is documented in section \ref{sec:data-description}.} Figure \ref{fig:stateSurvival} shows the survival probability (case ongoing) for the \gls{ni} cases for different States. As per our estimate, less than 20\% of the cases get disposed of within the prescribed timeline of 180 days. Also, the probability of case completion within 500 days is less than 50\% for most States.

\begin{figure}[!ht]
\centering
\caption{Survival probability of cases across States}\label{fig:stateSurvival}
\footnotesize
\textit{Note:} The X-axis shows the number of days that a case takes from the date of admission until the outcome. The Y-axis shows the probability of the case continuing up to a given number of days.
\includegraphics[width=\textwidth]{surv_states-1.png}
\end{figure}

Since cheque dishonour cases are a large burden on the judiciary, various committees, institutions, and government bodies have focused on expediting disposal. For example, the \cite{lci2008_138}, by relying on newspaper reports concerning the proportion of cheque dishonour cases, recommended setting up Fast Track Magisterial Courts (for a detailed discussion, refer to \cite{bhan2015_placing}). The Act was also subject to amendments in 2015 and 2018 to deal with the various procedural matters. Similarly, the Supreme Court has issued directions to lower courts and various financial institutions to take a more pragmatic and realistic approach for the speedy disposal of \gls{ni} cases.

In 2020, the Supreme Court of India took on board a suo-motu case concerning the “expeditious trial of cases under section 138 of the Negotiable Instruments Act 1881” \autocite{sc2020_138}. Among other directions, the court formed a \gls{coe}\footnote{Headed by Hon’ble Mr. Justice RC Chavan, former Judge of the Bombay High Court.} and appointed Amici Curiae\footnote{Mr. Sidharth Luthra (Sr Advocate) and Mr. K Parameshwar (Advocate).} to assist the court, to study processes to expedite the disposal of complaints under \S~138 of the \gls{ni}. In late 2020, the Amici submitted their recommendations. This included (i) increasing the use of pre-and post-summons mediation, (ii) expediting the service of summons to reduce absconsion, (iii) addressing the multiplicity of proceedings, etc. (for additional detail, refer to \cite{amicus2020_submission}).

Given this background, the purpose of this paper is fourfold. First, using case level eCourts data, for the first time in India, we estimate the real judicial caseload due to the \gls{ni}. Second, we estimate the average case disposal time and its variation over time and across States. Third, we build a conceptual framework to establish the link between specific case characteristics and procedural delay. Furthermore, we empirically test our conceptual framework of procedural determinants of case disposal time using regression analysis. In line with \cite{bielen2015}, using case level interim and final orders data, we establish the link between the specific case characteristics and procedural issues while controlling for inherent court and judge characteristics. The empirical literature on the length of court proceedings currently focuses mainly on determinants at the country and court levels, e.g. public budgets for courts, backlogs, number of judges, etc. Some studies also focus on the case level determinants like the number of parties involved, number of pleadings, etc. Given access to the interim and final orders data, we focus on the case characteristics that can be directly targeted using procedural reforms.

Additionally, we examine the role of the Negotiable Instruments (Amendment) Bill, 2015, which was passed by the Parliament in December 2015. The amendment focused on clarifying the jurisdiction-related issues for filing cases for an offence committed under \S~138 of the Negotiable Instruments Act, 1881. It also mandated the centralisation of cases against the same drawer. These reforms were specially drafted to accelerate court proceedings even though no evidence-based insight into the causes of prolonged trials was available. We analyse the effect of the 2015 amendment by comparing the duration of cases handled before and after their implementation using regression analysis. Our research makes it possible to evaluate the effectiveness of these reforms, at least for the States for which we gathered the data.

Our estimates suggest that cheque dishonour cases represent 13.2\% of courts' workload (pending and disposed) in India. This number could be underestimated because, in about 36\% of the cases in the sample, orders were either unavailable or could not be parsed. We also find significant variation in case disposal time across states. Median disposal time varies from as low as 218 days in Karnataka to as high as 547 days in Himachal Pradesh, with a median disposal time of 395 days. We also find that only a fifth of the cases gets disposed of within the prescribed limit of six months.

Controlling for State and year fixed effects, our findings suggest that, all else being equal, cases where the accused fails to appear take an additional 201 days and 7 hearings to dispose of. Similarly, cases run as summons trials take 112 days and 7.4 hearings more than summary trials. Cases with jurisdictional issues take 271 days and 5.5 hearings more to dispose of than cases without these issues. Cases with a multiplicity of proceedings take an additional 168 days and 9.9 hearings to dispose of than those without. Contested cases take fewer days to dispose of than uncontested cases, even though they take more hearings. Our results also indicate that the 2015 amendments have significantly reduced case duration -- reducing the case duration by more than six months.

The rest of this paper is organised as follows: after this introduction, \cref{sec:history} elaborates on the history of cheque dishonour provisions in India and their relation to judicial delays. In \cref{sec:select-case-char}, based on our conversations with practising lawyers, other stakeholders, and the recommendations of the Amici Curiae and the Supreme Court, we figure out the key case characteristics linked to the procedural delay that can be directly targeted by court interventions. \Cref{sec:methodology} describes the methodology, and \cref{sec:results} presents the results thereof. In \cref{sec:2015amend}, we analyse the effectiveness of the 2015 amendments by comparing the duration of cases handled before and after the procedural reforms were implemented. Finally, \cref{sec:conclusion} concludes the paper and presents the way forward.
